---
title: "Homework 1 - Monte Carlo Methods"
author: "Renjie Wei  UNI: rw2844"

output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message = FALSE)
```

# Problem 1
The standard Laplace distribution has density $f(x) = 0.5e^{-|x|}, x\in
(-\infty, \infty)$. Please provide an algorithm that uses the inverse transformation method to generate a random sample from this distribution. Use the $U(0,1)$ random number generator in  \em{\bf{R}}, write a  \em{\bf{R}}-function to implement the algorithm. Use visualization tools to validate your algorithm (i.e., illustrate whether the random numbers generated from your function truly follows the standard Laplace distribution.)

# Answer: 

Since $f(x) = 0.5e^{-|x|}, x\in
(-\infty, \infty)$.

Therefore,
$$F(x) = \int_{-\infty}^{x}0.5e^{-|s|}ds=\\
\int_{-\infty}^{x}0.5e^{s}ds\space(\text{if}\space x<0)\\
\int_{-\infty}^{0}0.5e^{s}ds+\int_{0}^{x}0.5e^{-s}ds\space(\text{if}\space x\ge0)$$
Hence,
$$F(x)= 0.5e^x\cdot \mathbb{I}(X<0)+(1-0.5e^{-x})\cdot \mathbb{I}(X\ge0)\\
=0.5[1+(-1)^{\mathbb{I}(X<0)} (1-e^{-|x|})]$$

To get the inverse function $F^{-1}(U)$, solve $u = F(x)$:
$$u=0.5e^{x} \ (\text{when}\space x<0\rightarrow 0\le u<0.5)\\
x=\ln( 2u)\\
or\\
u=1-0.5e^{-x} \ (\text{when}\space x\ge0\rightarrow 0.5\le u\le 1)\\
x=-\ln(2-2u)$$

```{r Problem_1_solution}
#Your R codes/functions
#install.packages("ExtDist")
library(ExtDist)
library(ggplot2)
set.seed(2022)
n=1e4
U<-runif(n)
L = rLaplace(n)
# inverse CDF
X = (U<0.5)*log(2*U)-(U>=0.5)*log(2-2*U)
dfL<-data.frame(
  var = c(rep("Laplace",n),rep("invCDF_sample",n)),
  value = c(L,X)
)
# the histogram of inverse CDF samples and samples from Laplace distribution
ggplot(dfL, aes(x=value, fill=var)) +
  geom_histogram(alpha=0.6, position='identity') +
  ggtitle("Histogram of samples and Laplcae distribution")
```

# Problem 2

Use the inverse transformation method to  derive an algorithm for generating a Pareto random number with $U\sim U(0,1)$, where the Pareto random number has a probability density function 
     $$f(x; \alpha, \gamma)=\frac{\gamma\alpha^{\gamma}}{x^{\gamma+1}} I\{x\ge \alpha\}$$ with two parameters $\alpha >0$ and $\gamma>0$. Use visualization tools to validate your algorithm (i.e., illustrate whether the random numbers generated from your function truly follows the target distribution.)


# Answer: 

Since $f(x; \alpha, \gamma)=\frac{\gamma\alpha^{\gamma}}{x^{\gamma+1}} I\{x\ge \alpha\},\alpha >0,\ \gamma>0$.

Therefore,
$$F(x)=\int_{\alpha}^{x}\frac{\gamma\alpha^{\gamma}}{s^{\gamma+1}}ds = \alpha^\gamma s^{-\gamma}\bigg|_{\alpha}^{x}=1-\alpha^\gamma x^{-\gamma}$$
To get the inverse function $F^{-1}(U)$, solve $u = F(x)$:
$$
u=1-\alpha^\gamma x^{-\gamma}(0<u\le1)\\
x = \frac{\alpha}{\sqrt[\gamma]{1-u}}
$$
```{r problem_2_solution}
#Your R codes/functions 
#install.packages("EnvStats")
library("EnvStats")
set.seed(2022)
n=1e4
Up<-runif(n)
# generating N random samples directly from Pareto(2,20) (parameters are tuned in order to make the stacked histogram readable)
alphaP <- 2
gammaP <- 20
P = rpareto(n,alphaP,gammaP)
# now using the inverse CDF method
Xp = 2/(1-Up)^(1/gammaP)

dfP<-data.frame(
  var = c(rep("Pareto",n),rep("invCDF_sample",n)),
  value = c(P,Xp)
)
# the histogram of inverse CDF samples and samples from Laplace distribution
ggplot(dfP, aes(x=value, fill=var)) +
  geom_histogram(alpha=0.6, position='identity') +
  ggtitle("Histogram of samples and Pareto distribution")
```

# Problem 3

Construct an algorithm for using the acceptance/rejection method to generate 100 pseudorandom
 variable from the pdf
 $$f(x) = \frac{2}{\pi \beta^2} \sqrt{\beta^2-x^2}, \,\, -\beta \le x \le \beta.$$
The simplest choice for $g(x)$ is the $U(-\beta, \beta)$ distribution but other choices are possible as well. Use visualization tools to validate your algorithm (i.e., illustrate whether the random numbers generated from your function truly follows the target distribution.)

# Answer: 
Since$f(x) = \frac{2}{\pi \beta^2} \sqrt{\beta^2-x^2}, \,\, -\beta \le x \le \beta$.

We can show that $$\max f(x) = \frac{2}{\pi}$$
When we decided to use the $U(-\beta, \beta)$ as the convenient distribution $g(x)$, then we should choose $M$ base on $M= \sup_{x\in[-\beta,\beta]} f(x)/g(x)=4\beta/\pi$.
```{r problem_3_solution}
#Your R codes/functions 
# First, create a function to calculate the pdf of x with input parameter beta and Uniform samples xs
targetdens <- function(beta,x){
  return((2/pi*(beta^2))*((beta^2-x^2)^1/2))
}

convdens <- function(beta,x){
  return(runif(length(x), min = -beta, max = beta))
}
# And the accept-rejection algorithm
myAccrej <- function(M, x, beta){
  U = runif(length(x))
  selected = x[U <= (targetdens(beta,x) / (M * convdens(beta,x)))]
  return(selected)
}

# Start sampling
set.seed(2022)
Beta = 1
M = (4*Beta)/pi
x = runif(1e6,-Beta,Beta)
samp = myAccrej(M=M,x,Beta)

# plot the sample histogram and the density curve, the plot shows that the shape of these two distributions are very close
hist(samp, prob=T, col="lightblue",xlab = "samples" )
curve((2/pi)*sqrt(1-x^2), -1,1, add=T, col="red")
```





